---
title: "Raincloud Plots with ggplot2" 
author:
    - name: "Cédric Scherer"
      url: https://cedricscherer.com
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
        highlight: kate
        code_folding: false  
        toc: true            
        toc_depth: 3        
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 8, fig.height = 6, dpi = 500)
```

# Packages

```{r}
library(ggplot2)
library(ggforce)
library(ggdist)
library(gghalves)

theme_set(theme_light(base_size = 16))
```

# Plots

## Violin-boxplot-combination

```{r}
g1 <- 
  ggplot(iris, aes(Species, Sepal.Width)) + 
  geom_violin(fill = "grey90") + 
  geom_boxplot(width = .2, outlier.shape = NA, coef = 0)

g1
```

## Better: add raw data 

### as jitter

```{r}
g1 + geom_point(alpha = .7, position = position_jitter(seed = 1))
```

... does look messy. One can play around with width but likely that it will never look clean.

```{r}
g1 + geom_point(alpha = .7, width = .1, position = position_jitter(seed = 1))
```


### as sina 

A sina ensures that the points are placed inside the violin. The [ggforce](https://ggforce.data-imaginist.com/) package comes with a `geom_sina()` function.

```{r}
g1 + ggforce::geom_sina(method = "counts", alpha = .5)
```

### as quasirandom distribution

Similarly, one can use the quasirandom geom from the [ggbeeswarm](https://github.com/eclarke/ggbeeswarm) package to reduce overplotting within categories:

```{r}
g1 + ggbeeswarm::geom_quasirandom(width = .3, alpha = .5, varwidth = TRUE, size = 2)
```


### as beeswarm

... or true beeswarms from the same package—*but note that it bins the data*.

```{r}
g1 + ggbeeswarm::geom_beeswarm(width = .3, alpha = .5, cex = 1.2, size = 2)
```


## Raincloud plots

We can combine a (half) violin with a summary plot (usually parts of a boxplot) and at thesame time show the raw data. This chart is called [raincloud plot](https://wellcomeopenresearch.org/articles/4-63).

### with linerange + dotplot

It is straightforard to build a verison of a raincloud plot with the [ggdist]() package. Here we use the `stat_halfeye` fucntion that draws a (half) violin and an interval slab to highlight important summary stats. 
I like the appearance a lot but *note that the errorbar style might be confusing* (and it was actually build to display confidence intervals).

We can combine this geom, scaled to show median, IQR and full range of the data (via `.width`), with `stat_dots` that draws a dot plot showing the raw data.

```{r}
ggplot(iris, aes(Species, Sepal.Width)) + 
  ggdist::stat_halfeye(adjust = .5, width = .3, .width = c(0.5, 1)) + 
  ggdist::stat_dots(side = "left", dotsize = .4, justification = 1.05, binwidth = .1)
```

### with boxplot + dotplot

We can replace the interval slab by a boxplot. We add some justification to the halfeye as well so the boxplot gets some space to fit in-between the **cloud** and the **rain**.

```{r}
ggplot(iris, aes(Species, Sepal.Width)) + 
  ggdist::stat_halfeye(adjust = .5, width = .3, .width = 0, justification = -.3, point_colour = NA) + 
  geom_boxplot(width = .1, outlier.shape = NA) +
  ggdist::stat_dots(side = "left", dotsize = .3, justification = 1.1, binwidth = .1)
```

(You can also use `gghalves::geom_half_dotplot(stackdir = "down")` to draw the dotplot.)

### with boxplot + jitter (on top)

Of course, if you don't like the binning of the data points, one can use a classical jitter.

```{r}
ggplot(iris, aes(Species, Sepal.Width)) + 
  ggdist::stat_halfeye(adjust = .5, width = .7, .width = 0, justification = -.2, point_colour = NA) + 
  geom_boxplot(width = .2, outlier.shape = NA) + 
  geom_jitter(width = .05, alpha = .3)
```

However, (as far as I know) there is no way to justify the jitter so it's shown next to the boxplot.

### with boxplot + jitter (side by side)

We can plot a side-by-side combination of jitter and boxplot with help of the [gghalves](https://erocoar.github.io/gghalves/) package.

```{r}
ggplot(iris, aes(Species, Sepal.Width)) + 
  ggdist::stat_halfeye(adjust = .5, width = .3, .width = 0, justification = -.3, point_colour = NA) + 
  geom_boxplot(width = .1, outlier.shape = NA) +
  gghalves::geom_half_point(side = "l", range_scale = .4, alpha = .5)
```

### with boxplot + barcode (side by side)

By repalcing the shape and restricting the range of the jitter, we can also turn it into a barcode plot:

```{r}
ggplot(iris, aes(Species, Sepal.Width)) + 
  ggdist::stat_halfeye(adjust = .5, width = .3, .width = 0, justification = -.3, point_colour = NA) + 
  geom_boxplot(width = .1, outlier.shape = NA) +
  gghalves::geom_half_point(side = "l", range_scale = 0, shape = 95, size = 15, alpha = .3)
```

***

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
git2r::repository()
sessionInfo()
```

</details>

